Interactive Visual Analysis of Lumbar Back Pain \subtitle{What the Lumbar Spine Tells About Your Life}

### 2014-09-29 draft.tex
First revised Version from Henry

% \abstract{The abstract should summarize the contents of the paper and should contain at least 70 and at most 200 words.}
\abstract{ 
The goal of epidemiology is to provide insight into disease causations.
%%
Hence, subject groups (cohorts) are analyzed to correlate the subjects' varying lifestyles, their medical properties and diseases.
%%
Recently, these cohort studies comprise medical image data.
%%
We assess the predictive quality of image-derived variables for lower back pain.
%%
Therefore, an Interactive Visual Analysis (IVA) framework was created and tested with 2,540 segmented lumbar spine data sets.
%%
The segmentation results are evaluated and quantified by employing shape-describing variables, such as the spine canal course, its curvature and torsion.
%%
We analyze the predictive power of shape-describing variables for non-image variables, e.g. pain indicators.
%%
Therefore, we automatically extract a decision tree classifier for each non-image variable.
%%
Furthermore, we provide an IVA technique to compare the classifiers with interactive scatterplot visualizations.
%%
As a first result, we conclude that image-based variables are only sufficient for classification of lifestyle factors.
%%
A correlation between lumbar spine shape and lower back pain disease was not existent in the automatically trained classifiers.
%%
However, the presented approach is a valuable extension for the IVA of epidemiological data.
%%
Hence, correlations between non-image variables were successfully detected and characterized.
}

@@

\section{\uppercase{Introduction}}
\label{sec:Introduction}
%%Epidemiology is the study of causation of diseases.
\noindent Epidemiology is the study of dissemination, causes and results of health-related states and events.
%%
For this purpose, large population studies, such as the \emph{Study of Health in Pomerania} (SHIP) \cite{SHIP}, gather as much information as possible about participants to be assessed towards different diseases.
%%
These information are used to determine risk factors for diseases, help people to live healthier or to support the diagnosis of widespread diseases.
%%
So far, epidemiological research is strongly hypothesis-driven.
%%
More precisely, observations made by clinicians are translated into hypotheses, which are then statistically evaluated using data variables from epidemiological studies.
%%
As a result, the hypothesis will be accepted or rejected.
%%

%For epidemiological research, cohort studies are acquired.
Cohort studies are an instrument of epidemiological research.
%For epidemiological research, cohort studies are carried out.
%%
Nowadays, these studies often comprise medical image data, e.g., tomographic image data sets, as well as general patient-specific attributes, such as \emph{gender}, \emph{age} and \emph{weight}.
%%
Back pain is one of the most frequent diseases in the western civilization and gains importance due to the population aging \cite{Hoy2010}.
%%
Epidemiological analysis of back pain aims to characterize the disease by determining risk factors.
%Many epidemiological-motivated approaches aim at improved characterization of this disease with respect to patient-specific attributes.
%Since back pain is one of the most frequent diseases in the western civilization, many epidemiological-motivated approaches aim at improved characterization of this disease with respect to patient-specific attributes.
%%
Although the shape and constitution of the spine, especially the lumbar spine, plays an important role for the presence of back pain, an automatic classification approach for characterization of back pain based on lumbar back spine attributes is still missing.
%%
Furthermore, previous analyses mainly focused on a single variable (e.g. \emph{BMI}) or a small subset of variables.
%%

We present a thorough analysis of the image-derived data of a cohort study including lumbar spine datasets.
%%
We analyze their discriminative power for different back pain characteristics.
%%
For this purpose, we combine data mining algorithms with data visualization techniques.
%%
%Then, interactive visual analysis provides insight into the quality of the image-derived data as well as their potential for possible back pain risk factors.
Then, Interactive Visual Analysis (IVA) provides insight into the quality of the image-derived data as well as their potential for being back pain risk factors.
%%
The workflow is part of an IVA framework, constructed to enhance the epidemiological workflow with iterative data driven techniques.
%%
Our contributions are:
\begin{itemize}
\item An IVA workflow for back pain analysis based on image-derived variables of 2,240 subjects,
%%
\item The identification and evaluation of lumbar spine shape properties for back pain diagnosis,
%%
\item The detection of correlations between image-based and socio-demographic as well as medical variables for an improved hypothesis generation,
%%
\item The identification of the most important variables via data mining methods (including a novel semi-quantitative evaluation concept) embedded in our IVA-framework
%%
\end{itemize}
%%
The analysis, as well as the presented IVA tool is available as supplementary material in a interactive \texttt{R Markdown}\footnote{Developed by RStudio, Inc; \texttt{rmarkdown.rstudio.com}} document (\url{blind.dnsalias.com}).
%%
%The IVA tool and all shown plots presented in this paper, except of Fig.~\ref{fig:age-gender}, are available as supplementary material under the following link: \url{blind.dnsalias.com}.
%% ---------------------------------------------
\section{\uppercase{Epidemiological Background}}
\label{sec:EpidemiologicalBackground}
%%
\noindent Epidemiological reasoning relies on a strict statistically-driven workflow \cite{Fletcher}:
%%
\begin{itemize}
%%
	\item Physicians formulate hypotheses based on observations made in their clinical practice.
%%
	\item Epidemiologists compile a list of variables depicting a hypothesis to prove it.
%%
	\item Statistical methods, such as regression analysis, assess the association of selected variables with the investigated disease.
%%
\end{itemize}
%%
Mutually dependent variables make this analysis challenging.
%%
Many diseases, such as different cancer types, are more likely with increasing age \cite{Fletcher}.
%%
For example, the effect of nutrition on prostate cancer requires age-normalization due to the increased incidence for older subjects.
%%
%When for example analyzing influences of nutrition to prostate cancer, the results need to be age-normalized.
%%
Age acts as a \emph{confounder} for prostate cancer.
%%
%These \emph{confounding} variables are often hard to find.
%%
Statistical correlation does not imply causation--epidemiologists need to assess the medical soundness of the statistical results.
%%------------------------------
%\subsection{Epidemiological Data}
\subsection{Challenges of Epidemiological Data Analysis}
Epidemiological data originates from a wide range of studies.
%%
The study type depends on the condition of interest.
%%
Most common are case-control studies analyzing one specific disease and its influence on the human body.
%%
We focus on large scale cohort study data.
%%
These studies aim at collecting as much data as possible for each subject.
%%
The collected data can be analyzed regarding many diseases and conditions.
%%

Epidemiological data are heterogenous and incomplete.
%%
For example, women-specific questions or data about a disease treatment only affects subjects suffering from this condition.
%%
Therefore, statistical analysis has to take missing data into account.
%%

Epidemiological data acquisition includes a variety of techniques, such as medical examinations, self reported questionnaires or genetic examinations.
%Epidemiologists acquire data using a wide range of techniques, such as medical examinations, self reported questionnaires or genetic examinations.
%%
This yields a heterogenous information space.
%%
To compare these data, information reduction techniques are necessary.
%%
For example, continuous data, such as age, is often discretized into quantile bins.
%%
%Every information reduction can introduce a bias to the data, since it proposes an assumption about the data.
Every information reduction proposes an assumption about the data and can therefore introduce a bias.
%%
For example, using age to divide subjects into \emph{young} and \emph{old} categories can distort statistical results.
%%
This distortion is reduced with increasing number of discretization steps.
%%

Modern cohort studies often comprise medical image data.
%%
The analysis of these data is challenging.
%These data are hard to analyze.
%%
Individual diagnosis or manual segmentation of each body structure by radiologists is tedious, costly and comprises little reproducibility.
%%
Segmentation algorithms are not generally available and need to be carefully adapted for each body structure.
%%
Segmentation data is usually analyzed by abstracting it into key figures, such as diameters or distances.
%%
These numeric values can be compared with non-image variables to retrieve correlations.
%%--------------------------------
\subsection{Lower Back Pain}
%%
%The lower (\emph{lumbar}) spine is the most stressed spine part .
%%
Lower back pain is one of the most frequent diseases in the western civilization \cite{Hoy2010}.
%%
Epidemiologists assume associations between lumbar back pain and lifestyle factors.
%%
These include nutrition, physical activity and body posture combined with physical stress at work.
%%
The exact causes as well as particularly vulnerable risk groups are unknown.
%%
Potential \emph{confounding} effects are also subject of current research.
%%

Epidemiologists want to characterize the healthy aging process of the spine.

@@

\cite{Pinheiro} applied association rule mining to create high-confidence decision rules based on variables, such as age, gender and geographic location about the mortality rate in liver cancer data.
%%
\cite{Sekhavat} propose an interactive visualization technique for decision trees, where all decisions are displayed in a heat map.
%%
The visualization translates the decision tree into a heat map, where each cell represents the connectivity between an initial and a target value.
%%
The resulting interactive visualization is well suited for analyzing single, but not multiple decision trees.
%%
\paragraph{SHIP Data Analysis.}
\cite{Klemm2013VMV} analyzed the lumbar spine variability of 490 \texttt{SHIP-2} subjects.
%%
They incorporate hierarchical agglomerative clustering to derive shape groups, yielding several groups of average shape and several outlier clusters.
%%
The extracted lumbar spine shape was correlated with subject size.
%%
Our data extraction process is based on the detection algorithm and lumbar spine canal extraction presented by \cite{Klemm2013VMV}.
%%
Clustering techniques applied to non-image data are strongly dependent on the chosen variables and distance measures \cite{Klemm2014BVM}.
%%
%Klemm et al. found \texttt{k-Prototypes} and \texttt{DBSCAN} appropriate for this task.
%%
\\\\
Unique for our work is the combination of data mining techniques with an IVA-approach by observing interesting feature relations in the context of image-derived features using multiple decision trees for one visualization.
%%
We abstract the decision tree results similar to \cite{Turkay}, making them comparable in an overview visualization.
%%
\section{\uppercase{The Lumbar Spine Data Set}}
\label{sec:MaterialsAndMethod}
%%
\noindent Our approach allows the simultaneous analysis of many variables.
%%
Therefore, epidemiologists compiled the data set with a wide range of variables possibly correlating with lumbar back pain.
%%
The data set comprises 6,753 subjects from two cohorts (4,420 from \texttt{SHIP-Trend-0} and 2,333 from \texttt{SHIP-2}) including non-image and image-derived data.
%%
%It is an active research area to determine these correlations.
%%

@@

\noindent Spine shape is confounded by several somatometric variables.

@@

Since men are on average taller than women and people of old age shrink due to bone erosion, gender and age are also confounders (Fig.~\ref{fig:age-gender}).

@@

All these variables need to be taken into account, when spine curvature and torsion is correlated with non-image variables.

@@

Since the gender confounder mainly encodes body size, we decided to divide subjects into body size groups.

@@

To avoid small outlier groups, epidemiologists recommend using quartiles to discretize metric variables.

@@

\section{\uppercase{Evaluation of Decision Trees}}
\label{sec:DecisionTrees}
%%
\noindent As described before, correlation coefficients fail to infer back pain status based on lumbar spine canal \emph{curvature} and \emph{torsion}.
%%
We rely on predictive classification trained to obtain a complex rule set on how combinations of the image-variable explain non-image variables.
%%
Decision trees are a popular classification method in data mining to create predictive models.
%%
%Leafs of a decision tree represent class labels, branches represent decision trees.
Leafs of a decision tree represent class labels, branches represent feature conjunctions leading to the class labels.
%%
Decision trees are easy to understand and to read.
%%
They work with numerical as well as categorical data.
%%
Epidemiologists can interpret the results without having deep knowledge about the classification algorithm.
%%
Readability is only granted for small trees. %, complex structures with many branches are not desirable.
%%
Too many branches also impose overfitting to the data \cite{DecisionTree}.

@@

Many follow-up variables are sparse, such as \emph{medication of diabetes} or \emph{reason of early retirement}.
%%
The classification algorithm may produce higher accuracy for variables with less subjects due to the small sample size.
%%
This makes these results less reliable.
%%
%We removed variables with less than 100 subjects, since they have too few subjects to be statistically significant.
Therefore we provide a way to adjust the minimal number of subjects for each variable using a slider input.
%%
The initial value is empirically set to $100$, marking a good tradeoff between sparse variables and statistical informative value.
%%
Furthermore, we map the number of subjects associated with a variable to point diameter in the scatter plot.
%%
This allows instant reliability assessment of the result.
%%

@@

\subsection{Size Groups}
%%
Many previously described results are confounded by \emph{subject size}.
%%
Differences between \emph{male} and \emph{female} subjects can be explained by the average \emph{body size} difference.
%%
For example, large subjects are already characterized by their rather straight spine.
%
The question is, whether the inter-group spine variability parameter is enough for predicting other variables or not.
%%
Dividing subjects into \emph{body size} groups potentially highlights classifications not confounded by \emph{body size}.
%%
\paragraph{Large Decision Trees. }
\emph{Back pain}-associated variables can be explained for various \emph{size}-groups, but we could not extract universal rules.
%%
\emph{Radiating back pain} could be described with error of $0.39$ using 23 rules for subjects between \emph{$171-177~cm$ body size} using \emph{torsion} and \emph{mean curvature}.
%%
For subjects sized \emph{$177-202~cm$} the error increases to $0.47$ using 20 decision rules.
%%
There are several decision trees for laboratory values, e.g. \emph{alanine aminotransferase} value (relevant for diagnosis of liver or gallbladder illness) in the blood can be described with an error of $0.4$ (\emph{$139-164~cm$}) to $0.36$ (\emph{$164 - 171~cm$}).
%%
Similar values can be observed for \emph{cholesterol} or \emph{age}.
%%
Due to the large decision trees, these results are not usable and impose overfitting to the data.
%%
\paragraph{Small Decision Trees. }
%%
The dummy variables show several variables described using only one decision rule with an error between $0.42$ to $0.47$.
%%
Most of these variables have a dominant manifestation and the classifier shows a low detection precision for the second manifestation.
%%
These variables include \emph{nutrition}, \emph{thyroid disorder} and \emph{social problems induced by back pain}.

@@

>>>>

% \abstract{The abstract should summarize the contents of the paper and should contain at least 70 and at most 200 words.}
\abstract{ 
The goal of epidemiology is to provide insight into disease causations.
%%
Hence, subject groups (cohorts) are analyzed to correlate the subjects' varying lifestyles, their medical properties and diseases.
%%
Recently, these cohort studies comprise medical image data.
%%
%We assess the predictive quality of image-derived variables for lower back pain.
We assess the relation between image-derived variables of the lumbar spine with lower back pain in a cross-sectional study.
%We assess the predictive quality of image-derived variables for lower back pain.
%%
Therefore, an Interactive Visual Analysis (IVA) framework was created and tested with 2,540 segmented lumbar spine data sets.
%%
The segmentation results are evaluated and quantified by employing shape-describing variables, such as the spine canal course, its curvature and torsion.
%%
%We analyze the predictive power of shape-describing variables for non-image variables, e.g. pain indicators.
We analyze mutual dependencies among shape-describing variables and non-image variables, e.g. pain indicators.
%%
Therefore, we automatically extract a decision tree classifier for each non-image variable.
%%
Furthermore, we provide an IVA technique to compare the classifiers with interactive scatterplot visualizations.
%%
As a first result, we conclude that image-based variables are only sufficient for classification of lifestyle factors within the data set.
%%
A correlation between lumbar spine shape and lower back pain was not existent in the automatically trained classifiers.
%%
However, the presented approach is a valuable extension for the IVA of epidemiological data.
%%
Hence, associations between non-image variables were successfully detected and characterized.
}

@@

\section{\uppercase{Introduction}}
\label{sec:Introduction}
%%Epidemiology is the study of causation of diseases.
\noindent Epidemiology is the study of dissemination, causes and results of health-related states and events.
%%
For this purpose, large population studies, such as the \emph{Study of Health in Pomerania} (SHIP) \cite{SHIP}, gather as much information as possible about participants to be assessed towards different diseases.
%%
This information is used to determine risk factors for diseases, help people to live healthier or to support the diagnosis of widespread diseases.
%%
So far, epidemiological research is strongly hypothesis-driven.
%%
More precisely, observations made by clinicians are translated into hypotheses, which are then statistically evaluated using data variables from epidemiological studies.
%%
As a result, the hypothesis will be accepted or rejected.
%%

%For epidemiological research, cohort studies are acquired.
Cohort studies are an instrument of epidemiological research.
%For epidemiological research, cohort studies are carried out.
%%
Nowadays, these studies often comprise medical image data, e.g., tomographic image data sets, as well as general patient-specific attributes, such as \emph{gender}, \emph{age} and \emph{weight}.
%%
Back pain is one of the most frequent complaints in the western civilization and gains importance due to the population aging \cite{Hoy2010}.
%%
Epidemiological analysis of back pain aims to determine risk factors.
%Many epidemiological-motivated approaches aim at improved characterization of this disease with respect to patient-specific attributes.
%Since back pain is one of the most frequent diseases in the western civilization, many epidemiological-motivated approaches aim at improved characterization of this disease with respect to patient-specific attributes.
%%
Although the shape and constitution of the spine, especially the lumbar spine, plays an important role for back pain, an automatic classification approach for characterization of back pain based on lumbar back spine attributes is still missing.
%%
Furthermore, previous analyses mainly focused on a single variable (e.g. \emph{BMI}) or a small subset of variables.
%%

We present a analysis of the image-derived data of a cohort study including lumbar spine datasets.
%%
We extract associations between spine shape and back pain characteristics.
%We analyze their discriminative power for different back pain characteristics.
%%
For this purpose, we combine data mining algorithms with data visualization techniques.
%%
%Then, Interactive Visual Analysis (IVA) provides insight into the quality of the image-derived data as well as their potential as back pain risk factors.
Then, Interactive Visual Analysis (IVA) highlights mutual dependencies between image-derived data and back pain related variables.
%%
The workflow is part of an IVA framework, constructed to enhance the epidemiological workflow with iterative data driven techniques.
%%
Our contributions are:
\begin{itemize}
\item An IVA workflow for back pain analysis based on image-derived variables of 2,240 subjects,
%%
\item The identification of lumbar spine shape properties associated with back pain,
%\item The identification and evaluation of lumbar spine shape properties associated with back pain,
%%
\item The detection of associations between image-based, socio-demographic and medical variables for hypothesis generation,
%%
\item The identification of the most important variables via data mining methods (including a novel semi-quantitative evaluation concept) embedded in our IVA-framework
%%
\end{itemize}
%%
The analysis, as well as the presented IVA tool is available as supplementary material in a interactive \texttt{R Markdown}\footnote{Developed by RStudio, Inc; \texttt{rmarkdown.rstudio.com}} document (\url{blind.dnsalias.com}).
%%
%The IVA tool and all shown plots presented in this paper, except of Fig.~\ref{fig:age-gender}, are available as supplementary material under the following link: \url{blind.dnsalias.com}.
%% ---------------------------------------------
\section{\uppercase{Epidemiological Background}}
\label{sec:EpidemiologicalBackground}
%%
\noindent Epidemiological reasoning relies on a strict statistically-driven workflow \cite{Fletcher}:
%%
\begin{itemize}
%%
	\item Physicians formulate hypotheses based on observations made in their clinical practice.
%%
	\item Epidemiologists compile a list of variables depicting a hypothesis to prove it.
%%
	\item Statistical methods, such as regression analysis, assess the association of selected variables with the investigated disease.
%%
\end{itemize}
%%
Mutually dependent variables make this analysis challenging.
%%
Many diseases, such as different cancer types, are more likely with increasing age \cite{Fletcher}.
%%
For example, the effect of nutrition on prostate cancer requires age-normalization due to the increased incidence for older subjects.
%%
%When for example analyzing influences of nutrition to prostate cancer, the results need to be age-normalized.
%%
Age acts as a \emph{risk factor} for prostate cancer.
%%
%These \emph{confounding} variables are often hard to find.
%%
Statistical correlation does not imply causation--epidemiologists need to assess the medical soundness of the statistical results.
%%------------------------------
%\subsection{Epidemiological Data}
\subsection{Challenges of Epidemiological Data Analysis}
Epidemiological data originates from a wide range of studies.
%%
The study type depends on the condition of interest.
%%
Most common are case-control studies analyzing one specific disease and its influence on the human body.
%%
We focus on large scale cohort study data.
%%
These studies aim at collecting as much data as possible for each subject.
%%
The collected data can be analyzed regarding many diseases and conditions.
%%

Epidemiological data are heterogenous and incomplete.
%%
%For example, women-specific questions or data about a disease treatment only affects subjects suffering from this condition.
For example, data about a disease treatment only affects subjects suffering from this condition.
%%
Therefore, statistical analysis has to take missing data into account.
%%

Epidemiological data acquisition includes a variety of techniques, such as medical examinations, self reported questionnaires or genetic examinations.
%Epidemiologists acquire data using a wide range of techniques, such as medical examinations, self reported questionnaires or genetic examinations.
%%
This yields a heterogenous information space.
%%
To compare these data, information reduction techniques are necessary.
%%
For example, continuous data, such as age, is often discretized into quantile bins for visualizing age-dependent disease frequencies.
%%
%Every information reduction can introduce a bias to the data, since it proposes an assumption about the data.
Every information reduction proposes an assumption about the data and can therefore introduce a bias.
%%
For example, using age to divide subjects into \emph{young} and \emph{old} categories can distort statistical results.
%%
This distortion is reduced with increasing number of discretization steps.
%%

Modern cohort studies often comprise medical image data.
%%
The analysis of these data is challenging.
%These data are hard to analyze.
%%
Individual diagnosis or manual segmentation of each body structure by radiologists is tedious, costly and comprises little reproducibility.
%%
Segmentation algorithms are not generally available and need to be carefully adapted for each body structure.
%%
Segmentation data is usually analyzed by abstracting it into key figures, such as diameters or distances.
%%
These numeric values can be compared with non-image variables to retrieve correlations.
%%--------------------------------
\subsection{Lower Back Pain}
%%
%The lower (\emph{lumbar}) spine is the most stressed spine part .
%%
Lower back pain is one of the most frequent complaints in the western civilization \cite{Hoy2010}.
%%
Epidemiologists assume associations between lumbar back pain and lifestyle factors.
%%
These include nutrition, physical activity and body posture combined with physical stress at work.
%%
The exact causes as well as particularly vulnerable risk groups are unknown.
%%
%Potential \emph{confounding} effects are also subject of current research.
%%

%Epidemiologists want to characterize the healthy aging process of the spine.
Epidemiologists want to characterize the relation between the aging and degenerative process of the spine \cite{Szpalski2005}. %healthy aging process of the spine.

@@

\cite{Pinheiro} applied association rule mining to create high-confidence decision rules based on variables, such as age, gender and geographic location about the mortality rate in liver cancer data.
%%
\cite{Sekhavat} propose an interactive visualization technique for decision trees, where all decisions are displayed in a heat map.
%%
The visualization translates the decision tree into a heat map, where each cell represents the connectivity between an initial and a target value.
%%
The resulting interactive visualization is well suited for analyzing single, but not multiple decision trees.
%%
\paragraph{SHIP Data Analysis.}
\cite{Klemm2013VMV} analyzed the lumbar spine variability of 490 \texttt{SHIP-2} subjects.
%%
They incorporate hierarchical agglomerative clustering to derive shape groups, yielding several groups of average shape and several outlier clusters.
%%
The extracted lumbar spine shape was correlated with subject size.
%%
Our data extraction process is based on the detection algorithm and lumbar spine canal extraction presented by \cite{Klemm2013VMV}.
%%
%Clustering techniques applied to non-image data are strongly dependent on the chosen variables and distance measures \cite{Klemm2014BVM}.
Clustering techniques are strongly dependent on the chosen variables and distance measures \cite{Klemm2014BVM}.
%%
%Klemm et al. found \texttt{k-Prototypes} and \texttt{DBSCAN} appropriate for this task.
%%
\\\\
Unique for our work is the combination of data mining techniques with an IVA-approach by observing interesting feature relations in the context of image-derived features using multiple decision trees for one visualization.
%%
We abstract the decision tree results similar to \cite{Turkay}, making them comparable in an overview visualization.
%%
\section{\uppercase{The Lumbar Spine Data Set}}
\label{sec:MaterialsAndMethod}
%%
\noindent Our approach allows the simultaneous analysis of many variables.
%%
Therefore, epidemiologists compiled the data set with a wide range of variables possibly correlating with lumbar back pain.
%%
The data set comprises 6,753 subjects from two cohorts (4,420 from \texttt{SHIP-Trend-0} and 2,333 from \texttt{SHIP-2}) including non-image and image-derived data.
%%
%It is an active research area to determine these correlations.
%%

@@

\noindent Spine shape is possibly associated with several somatometric variables.

@@

Since men are on average taller than women and people of old age shrink due to bone erosion, gender and age are also risk factors (Fig.~\ref{fig:age-gender}).

@@

To assess their influence, we take them into account when spine curvature and torsion is correlated with non-image variables.
%All these variables need to be taken into account, when spine curvature and torsion is correlated with non-image variables.

@@

Since the gender encodes body size, we decided to divide subjects into body size groups.

@@

Discretizing metric variables using quartiles avoids small outlier groups.
%To avoid small outlier groups, epidemiologists recommend using quartiles to discretize metric variables.

@@

\section{\uppercase{Evaluation of Decision Trees}}
\label{sec:DecisionTrees}
%%
\noindent As described before, correlation coefficients fail to infer back pain status based on lumbar spine canal \emph{curvature} and \emph{torsion}.
%%
We rely on predictive classification trained to obtain a complex rule set on how combinations of the image-variable explain non-image variables.
%%
Decision trees are a popular classification method in data mining to create predictive models.
%%
%Leafs of a decision tree represent class labels, branches represent decision trees.
Leafs of a decision tree represent class labels, branches represent feature conjunctions leading to the class labels.
%%
Decision trees are easy to understand and to read.
%%
They work with numerical as well as categorical data.
%%
%Epidemiologists can interpret the results without having deep knowledge about the classification algorithm.
%%
Readability is only granted for small trees. %, complex structures with many branches are not desirable.
%%
Too many branches also impose overfitting to the data \cite{DecisionTree}.

@@

Many interdependent variables are sparse, such as \emph{medication of diabetes} or \emph{reason of early retirement}.
%%
The classification algorithm may produce higher accuracy for variables with less subjects due to the small sample size.
%%
This makes these results less reliable.
%%
%We removed variables with less than 100 subjects, since they have too few subjects to be statistically significant.
Therefore we provide a way to adjust the minimal number of subjects for each variable using a slider input.
%%
The initial value is empirically set to $100$, marking a good tradeoff between sparse variables and statistical informative value.
%%
Furthermore, we map the number of subjects associated with a variable to point diameter in the scatter plot.
%%
This allows instant reliability assessment of the result.
%%

@@

\subsection{Size Groups}
%%
Many previously described results are influenced by \emph{subject size}.
%%
Differences between \emph{male} and \emph{female} subjects can be explained by the average \emph{body size} difference.
%%
For example, large subjects are already characterized by their rather straight spine.
%
The question is, whether the inter-group spine variability parameter is enough for predicting other variables or not.
%%
Dividing subjects into \emph{body size} groups potentially highlights classification not influenced by \emph{body size}.
%%
%Dividing subjects into \emph{body size} groups potentially highlights classifications not confounded by \emph{body size}.
%%
\paragraph{Large Decision Trees. }
\emph{Back pain}-associated variables can be explained for various \emph{size}-groups, but we could not extract universal rules.
%%
\emph{Radiating back pain} could be described with error of $0.39$ using 23 rules for subjects between \emph{$171-177~cm$ body size} using \emph{torsion} and \emph{mean curvature}.
%%
For subjects sized \emph{$177-202~cm$} the error increases to $0.47$ using 20 decision rules.
%%
There are several decision trees for laboratory values, e.g. \emph{alanine aminotransferase} value (relevant for diagnosis of liver or gallbladder illness) in the blood can be described with an error of $0.4$ (\emph{$139-164~cm$}) to $0.36$ (\emph{$164 - 171~cm$}).
%%
Similar values can be observed for \emph{cholesterol} or \emph{age}.
%%
Due to the large decision trees, these results are not usable and impose overfitting to the data.
%%
\paragraph{Small Decision Trees. }
%%
The dummy variables show several variables described using only one decision rule with an error between $0.42$ to $0.47$.
%%
Most of these variables have a dominant manifestation and the classifier shows a low detection precision for the second manifestation.
%%
These variables include \emph{nutrition}, \emph{thyroid disorder} and \emph{social problems induced by back pain}.

@@

The methods presented herein may be applied to comprehensive epidemiological data sets to learn more about mutual dependencies among variables and to generate hypotheses on potential associations and subgroups.
%
These hypotheses, however, have to be substantiated by dedicated statistical analyses and replication in independent cohorts.
